---
title: "Untitled"
author: "Angela Threadgill"
date: "10/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load necessary libraries
library(tidyverse)
library(ggridges)
```

```{r}
# read data from GitHub
# https://fred.stlouisfed.org/series/CASANF0POP
# https://fred.stlouisfed.org/series/MHICA06075A052NCEN
# https://fred.stlouisfed.org/series/USACP040100GPM
# https://fred.stlouisfed.org/series/QUSN628BIS

rent_df <- 
  readr::read_csv(
    file = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-05/rent.csv"
  ) %>% 
  dplyr::select(
    -c(date, address, lat, lon, descr, details)
  ) %>% 
  dplyr::filter(
    county != "napa"
  )

new_constr_df <- 
  readr::read_csv(
    file = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-05/new_construction.csv"
  ) %>% 
  dplyr::select(
    -c(starts_with("the_geom"), source)
  )

files_list <- list.files(path = "data/")

income_df <-
  purrr::map2_dfr(
    .x = files_list[grepl(x = files_list, "income")],
    .y = rent_df %>% dplyr::filter(!is.na(county)) %>% distinct(county) %>% arrange(county) %>% pull(county),
    .f = ~ readr::read_csv(
      file = glue::glue("data/{.x}")
    ) %>% 
      dplyr::mutate(
        county = .y
      )
  ) %>% 
  dplyr::rename_all(
    tolower
  ) %>% dplyr::mutate(
    year = lubridate::year(as.Date(date)),
    income = as.numeric(income)
  ) %>% 
  dplyr::select(
    -date
  ) %>% 
  dplyr::filter(
    year >= 2000
  )
```

We notice that the counties in the `rent_df` dataset are all lowercase and do not have the word "county" appended to them.

```{r}
rent_df %>% 
  dplyr::distinct(
    county
  )
```
Additionally, we notice that the counties in `new_cunstr_df` are title case with the word "county" appended. To match these allowable values with that of `rent_df`, we can convert to lowercase and remove the word "county". We also observe that `rent_df` contains Santa Cruz and a null county while `new_constr_df` does not.

```{r}
new_constr_df %>% 
  dplyr::distinct(
    county
  )
```

Let's go ahead and use `tolower()` and `stringr::str_remove_all()` on `county` in `new_constr_df`.

```{r}
new_constr_clean_df <-
  new_constr_df %>% 
  dplyr::mutate(
    county = tolower(stringr::str_remove_all(county, "County"))
  )

new_constr_clean_df
```
We observe 1,394 entries with null counties. Given these comprise of 0.7% of the `rent_df` dataset, let's go ahead and filter those out.

```{r}
rent_df %>% 
  dplyr::count(
    county
  )
```

```{r}
rent_clean_df <-
  rent_df %>% 
  dplyr::filter(
    !is.na(county)
  )
```


Now, let's check if `rent_clean_df` is distinct by `post_id`. Let's also check if it's distinct by `title`, since individuals can create duplicate posts for the same apartment.

```{r}
rent_clean_df %>% 
  dplyr::summarise(
    n = n(),
    n_dist = n_distinct(post_id)
  )

rent_clean_df %>% 
  dplyr::summarise(
    n = n(),
    n_dist = n_distinct(title)
  )
```

We notice that `rent_clean_df` is distinct by `post_id`, but **not** by `title`.

```{r}
dup_titles <-
  rent_clean_df %>% 
  dplyr::count(
    title
  ) %>% 
  dplyr::filter(
    n > 1
  ) %>% 
  dplyr::pull(
    title
  )
```

We notice that there are ~25,000 entries that have duplicate titles. However, after some research, it looks as though certain individuals are posting the same title, but for a different unit. Here, a "different unit" is identified as the same title, but in a different neighborhood/county and with a different price and square footage.

```{r}
dup_df <-
  rent_clean_df %>% 
  dplyr::filter(
    title %in% dup_titles
  )

dup_df %>% 
  dplyr::count(
    nhood,
    county, 
    title,
    price,
    sqft
  ) %>% 
  dplyr::filter(
    n > 1
  )
```

```{r}
rent_clean_df <-
  rent_clean_df %>% 
  dplyr::distinct(
    nhood,
    county, 
    title,
    price,
    sqft,
    .keep_all = TRUE
  )
```

```{r}
rent_clean_df %>% 
  dplyr::group_by(
    year,
    county
  ) %>% 
  dplyr::summarise(
    avg_rent = mean(price, na.rm = TRUE)
  ) %>% 
  ggplot(
    aes(x = year, y = avg_rent, color = county)
  ) +
  geom_line() +
  theme_minimal()
```

```{r}
income_df %>% 
  ggplot(
    aes(x = year, y = income, color = county)
  ) +
  geom_line() +
  theme_minimal()
```

```{r}
avg_rent_df <-
  rent_clean_df %>% 
  dplyr::group_by(
    year,
    county
  ) %>% 
  dplyr::summarise(
    avg_rent = mean(price, na.rm = TRUE)
  )

avg_rent_df %>% 
  dplyr::inner_join(
    income_df,
    by = c("year", "county")
  ) %>% 
  dplyr::mutate(
    mth_income = income / 12,
    affordability = mth_income / avg_rent
  ) %>% 
  ggplot(
    aes(x = year, y = affordability, color = county)
  ) +
  geom_line() +
  theme_minimal() +
  geom_hline(
    yintercept = 3
  )
```

```{r}
# maybe pivot longer and plot both the income and rental price distributions
rent_clean_df %>%
  ggplot() +
  geom_density(
    aes(x = price, color = county)
  ) +
  scale_x_log10() +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    axis.title = element_blank(),
    axis.text.y = element_blank()
  ) +
  labs(
    title = "Distribution of rental prices across counties in the Bay Area"
  )
# 
# rent_clean_df %>%
#   ggplot() +
#   geom_density_ridges(
#     aes(x = price, y = county)
#   ) +
#   scale_x_log10() +
#   theme_minimal() +
#   theme(
#     panel.grid.major = element_blank(),
#     axis.title = element_blank(),
#     axis.text.y = element_blank()
#   )
```

```{r}
# maybe a metric like the proportion of years that a county is considered unafforadable under the affordability metric in this analysis
```

