---
title: "Exploratory data analysis (EDA)"
author: "Angela Threadgill"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(vars)
```

```{r}
rent_df <- 
  readr::read_csv(
    file = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-05/rent.csv"
  ) %>% 
  dplyr::select(
    -c(date, address, lat, lon, descr, details, city, baths, room_in_apt)
  )

files_list <- list.files(path = "data/")

income_df <-
  purrr::map2_dfr(
    .x = files_list[grepl(x = files_list, "income")],
    .y = rent_df %>% dplyr::filter(!is.na(county), county != "napa") %>% distinct(county) %>% arrange(county) %>% pull(county),
    .f = ~ readr::read_csv(
      file = glue::glue("data/{.x}")
    ) %>% 
      dplyr::mutate(
        county = .y
      )
  ) %>% 
  dplyr::rename_all(
    tolower
  ) %>% dplyr::mutate(
    year = lubridate::year(as.Date(date)),
    income = as.numeric(income)
  ) %>% 
  dplyr::select(
    -date
  ) %>% 
  dplyr::filter(
    year >= 2000
  )

pop_df <-
  purrr::map2_dfr(
    .x = files_list[grepl(x = files_list, "pop")],
    .y = rent_df %>% dplyr::filter(!is.na(county), county != "napa") %>% distinct(county) %>% arrange(county) %>% pull(county),
    .f = ~ readr::read_csv(
      file = glue::glue("data/{.x}")
    ) %>% 
      dplyr::mutate(
        county = .y
      )
  ) %>% 
  dplyr::rename_all(
    tolower
  ) %>% dplyr::mutate(
    year = lubridate::year(as.Date(date)),
    population = as.numeric(population)
  ) %>% 
  dplyr::select(
    -date
  ) %>% 
  dplyr::filter(
    year >= 2000
  )

# https://fred.stlouisfed.org/series/CPIAUCSL
cpi_df <-
  data.table::fread(
    file = "data/us-cpi.csv"
  ) %>% 
  janitor::clean_names(
    case = "snake"
  ) %>% 
  dplyr::mutate(
    year = lubridate::year(date)
  ) %>% 
  dplyr::group_by(
    year
  ) %>% 
  dplyr::summarise(
    cpi = mean(cpiaucsl)
  ) %>% 
  dplyr::filter(
    year >= 2000,
    year <= 2018
  )
```

## Data munging

```{r}
rent_clean_df <-
  rent_df %>% 
  dplyr::filter(
    !is.na(county)
  ) %>% 
  dplyr::distinct(
    nhood,
    county, 
    title,
    price,
    beds,
    sqft,
    .keep_all = TRUE
  )
```

## EDA

```{r}
rent_clean_df %>% 
  dplyr::count(
    year
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    prop = n / sum(n)
  ) %>% 
  ggplot(
    aes(x = year, y = prop)
  ) +
  geom_col(
    fill = "#5a7572"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank()
  ) +
  labs(
    title = "% of total Craigslist rental postings in the Bay Area from 2000 - 2018"
  ) +
  scale_y_continuous(
    labels = scales::percent
  )
```

```{r}
rent_clean_df %>% 
  dplyr::count(
    county
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    prop = n / sum(n)
  ) %>% 
  ggplot(
    aes(x = county, y = prop)
  ) +
  geom_col(
    fill = "#5a7572"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank()
  ) +
  labs(
    title = "% of total Craigslist rental postings in the Bay Area from 2000 - 2018"
  ) +
  scale_y_continuous(
    labels = scales::percent
  )
```

```{r}
rent_clean_df %>% 
  dplyr::count(
    year,
    county
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(
    county
  ) %>% 
  dplyr::mutate(
    prop = n / sum(n)
  ) %>% 
  ggplot(
    aes(x = year, y = prop)
  ) +
  geom_col(
    fill = "#5a7572"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    title = "% of total Craigslist rental postings in the Bay Area from 2000 - 2018 by county"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  facet_wrap(
    . ~ county
  )
```

When observing the underlying data, we find that 2000 - 2002 overall contain low volume and that Santa Cruz and Solano have multiple years where the are less than 50 rental postings. Given concerns around sufficient data for statistical inference and analysis within these counties and years, I am removing these counties from the analysis.

```{r}
rent_clean_df2 <-
  rent_clean_df %>% 
  dplyr::filter(
    !(county %in% c("santa cruz", "solano", "napa")),
    !(year %in% 2000:2002)
  )
```

This results in a 7% reduction in the dataset after having already data munged. The final record count is 183,872.

```{r}
rent_clean_df2 %>% 
  dplyr::select(
    post_id,
    price,
    beds,
    sqft
  ) %>% 
  tidyr::pivot_longer(
    cols = c(price, beds, sqft)
  ) %>% 
  ggplot(
    aes(x = value, color = name)
  ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  scale_x_log10() +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_blank()
  )
```

From the view above and table below, we can see that the data for each post does not seem unreasonably large or small. Given this, we can drop `beds` and `sqft` from the analysis since we no longer need them going forward.

```{r}
summary(rent_clean_df2)
```

```{r}
rent_clean_df3 <-
  rent_clean_df2 %>% 
  dplyr::select(
    -c(beds, sqft)
  )
```

```{r}
rent_clean_df3 %>% 
  ggplot(
    aes(x = price, color = county)
  ) +
  geom_density() +
  theme_minimal() +
  scale_x_log10() +
  labs(
    title = "Distribution of rental prices from 2003-2018 by county"
  ) +
  facet_wrap(
    . ~ year
  ) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank()
  )
```

From the visual above, we see that the distribution of rental prices by county begins to diverge after the financial crisis, especiall in years 2011 onward.

```{r}
rent_cpi_df <-
  rent_clean_df3 %>% 
  dplyr::group_by(
    county, year
  ) %>% 
  dplyr::summarise(
    avg_price = median(price, na.rm = TRUE)
  ) %>% 
  dplyr::inner_join(
    cpi_df,
    by = "year"
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(
    county
  ) %>% 
  dplyr::mutate(
    `Rental price index` = (avg_price / avg_price[1L]) * 100,
    cpi = (cpi / cpi[1L]) * 100
  ) %>% 
  dplyr::select(
    -avg_price
  ) %>% 
  dplyr::rename(
    CPI = cpi
  ) %>% 
  tidyr::pivot_longer(
    cols = c(`Rental price index`, CPI)
  ) 

rent_cpi_df %>% 
  ggplot(
    aes(x = year, y = value, color = name)
  ) +
  geom_line() +
  facet_wrap(
    . ~ county
  ) +
  theme_minimal() +
  scale_color_brewer(
    palette = "Set2"
  ) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank()
  ) +
  labs(
    title = "Rent price index and U.S. CPI overtime"
  )
```

```{r}
rent_pop_df <-
  rent_clean_df3 %>% 
  dplyr::group_by(
    county, 
    year
  ) %>% 
  dplyr::summarise(
    avg_price = mean(price, na.rm = TRUE)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::inner_join(
    pop_df,
    by = c("county", "year")
  ) %>% 
  dplyr::group_by(
    county
  ) %>% 
  dplyr::mutate(
    `YoY change in rent price` = (avg_price - lag(avg_price)) / lag(avg_price),
    `YoY change in population` = (population - lag(population)) / lag(population)
  ) %>% 
  tidyr::pivot_longer(
    cols = c(`YoY change in population`)
  )

rent_pop_df %>% 
  dplyr::filter(
    !is.na(value)
  ) %>% 
  ggplot(
    aes(x = year, y = value, color = name)
  ) +
  geom_line() +
  facet_wrap(
    . ~ county
  ) +
  theme_minimal() +
  scale_color_brewer(
    palette = "Set2"
  ) +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    panel.grid.major = element_blank()
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    title = "Year-over-year change in population across time"
  )
```
```{r}
rent_cpi_df2 <-
  rent_clean_df3 %>% 
  dplyr::group_by(
    county, year
  ) %>% 
  dplyr::summarise(
    avg_price = median(price, na.rm = TRUE)
  ) %>% 
  dplyr::inner_join(
    cpi_df,
    by = "year"
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(
    county
  ) %>% 
  dplyr::mutate(
    cpi = (cpi / cpi[1L])
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    inf_adj_price = avg_price / cpi
  ) %>% 
  dplyr::group_by(
    county
  ) %>% 
  dplyr::mutate(
    yoy_price = (inf_adj_price - lag(inf_adj_price)) / lag(inf_adj_price)
  ) %>% 
  tidyr::pivot_longer(
    cols = c(yoy_price)
  )

rent_cpi_df2 %>% 
  dplyr::filter(
    !is.na(value)
  ) %>% 
  ggplot(
    aes(x = year, y = value, color = name)
  ) +
  geom_line() +
  facet_wrap(
    . ~ county
  ) +
  theme_minimal() +
  scale_color_brewer(
    palette = "Set2"
  ) +
  theme(
    #legend.position = "none",
    axis.title = element_blank(),
    panel.grid.major = element_blank()
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    title = "Year-over-year change in inflation-adjusted median rental price across time"
  )
```

```{r}
main_df <-
  rent_clean_df3 %>% 
  dplyr::group_by(
    year,
    county
  ) %>% 
  dplyr::summarise(
    avg_rent = mean(price, na.rm = TRUE)
  ) %>% 
  dplyr::inner_join(
    income_df,
    by = c("year", "county")
  ) %>% 
  dplyr::mutate(
    mth_income = income / 12,
    affordability = mth_income / avg_rent
  ) %>% 
  dplyr::inner_join(
    pop_df,
    by = c("year", "county")
  ) %>% 
  dplyr::inner_join(
    cpi_df,
    by = c("year")
  )

main_df %>% 
  dplyr::filter(
    county == "san francisco"
  ) %>% 
  ggplot(
    aes(x = year, y = affordability, color = county)
  ) +
  geom_line() +
  theme_minimal() +
  geom_hline(
    yintercept = 3
  ) +
  scale_color_brewer(
    palette = "Set2"
  ) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  ) +
  labs(
    title = "Affordability metric across time",
    subtitle = "San Francisco county"
  )
```

```{r}
main_df2 <- main_df %>% dplyr::filter(county == "san francisco")

model1 = vars::VAR(main_df2[c("affordability"," population")], p = 2)
summary(model1)
```

